Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-17 paper/ms.tex
some words about alternative methods

\section{celerite as a model of stellar variations}

@@

To demonstrate this method, we will use a very simple heuristic model based on
\dfmtodo{CITE} where the model PSD is given by a mixture of 9 components with
amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
several nuisance parameters.
The first two terms are used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$ in each term.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be easily extended to include small frequency splitting and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by physical parameters
like $\log g$.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{figures/transit-corner.pdf}
\caption{The marginalized posterior constraints on the physical parameters of
    the planet transiting during the light curve shown in the top panel of
    \Figure{transit-ml}.
    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
    regions in the marginalized planes and the histograms along the diagonal
    show the marginalized posterior for each parameter.
    The true values used in the simulation are indicated by blue lines.
    For each parameter, the inference is consistent with the true value.
    \figurelabel{transit-corner}}
\end{center}
\end{figure}

@@

Toeplitz, KISS-GP, CARMA, HODLR.

Limitations of celerite: one-dimension, stationary, etc.

>>>>

\section{celerite as a model of stellar variations}\sectlabel{sho}

@@

To demonstrate this method, we will use a very simple heuristic model based on
\dfmtodo{CITE} where the model PSD is given by a mixture of 8 components with
amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be easily extended to include small frequency splitting and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by physical parameters
like $\log g$.

@@

\begin{figure}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/transit-corner.pdf}
\caption{The marginalized posterior constraints on the physical parameters of
    the planet transiting during the light curve shown in the top panel of
    \Figure{transit-ml}.
    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
    regions in the marginalized planes and the histograms along the diagonal
    show the marginalized posterior for each parameter.
    The true values used in the simulation are indicated by blue lines.
    For each parameter, the inference is consistent with the true value.
    \figurelabel{transit-corner}}
\end{center}
\end{figure}

@@

There are many other methods of scaling GP models to large datasets and in
this section we will explore some comparisons between \celerite\ and some
other popular methods and discuss some use cases where \celerite\ shines and
others where it does not.
Scalable GP methods tend to fall into two categories: approximate and
restrictive.
\celerite\ falls into the latter category because, while the method is exact,
it requires a specific choice of stationary kernel function and it can only be
used in one-dimension.

\dfmtodo{Add discussion of original R\&P method here.}

Another commonly used restrictive method uses the fact that, in the limit of
evenly-spaced data and homoscedastic uncertainties, the covariance matrix is
``Toeplitz'' \citep[for example][]{Dillon:2013}.
There are exact methods for solving Toeplitz matrix equations that scale as
$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N$
\citep{Wilson:2014}.
The Toeplitz method is, in some ways, more flexible than \celerite\ because it
can be used with any stationary kernel but it requires uniformly spaced data
and the scaling is worse than \celerite\ so it will, in general, be less
efficient when applied to large datasets.

The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
Like \celerite, the likelihood function for a CARMA model conditioned on a
sorted one-dimensional dataset can be solved in $\mathcal{O}(N)$ using a
recursive method.
The kernel function for CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to a \celerite\ and the parameters $a_j$, $b_j$,
$c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not true, however, for any $J > 1$.
This means that \celerite\ could be used to compute any CARMA model but using
CARMA to compute the likelihood of a \celerite\ model would require solving
\eq{carma-coeff} for a given set of $\{A_j\}_{j=1}^J$ numerically.
This is an interesting distinction because, while the computational scaling of
CARMA models is also $\mathcal{O}(N\,J^2)$, CARMA solvers are, in practice,
faster and the memory requirements are smaller.
However, as discussed in \sect{sho}, \celerite\ models have physical a
interpretation as a mixture of stochastically-driven damped harmonic
oscillators.
This enabled, for example, the use of \celerite\ as a model for asteroseismic
oscillations.
A similar example using CARMA would necessarily be qualitative.

Approximate:
- inducing point methods \citep{Quinonero-Candela:2007}
- KISS-GP \citep{Wilson:2015, Wilson:2015a}
- HODLR \citep{Ambikasaran:2016}
